# (PART) Introduction {-}

# Introduction to R

```{r 010.01, include=FALSE}
# Load some libraries
library(here)

knitr::opts_chunk$set(
  collapse = TRUE,
  strip.white = TRUE,
  comment = "#>",
  out.width = "65%",
  message=FALSE,
  warnings=FALSE
)

# Set the rproj home directory
here::here()

# Turn on the DataCamp Light interactive tools
tutorial::go_interactive()

```

Statistical analysis is the core for nearly all research projects and researchers have a wide variety of statistical tools that they can use, like _SPSS_ and _SAS_. Unfortunately, these analysis tools are expensive or difficult to master so this lab manual introduces _R_, a powerful, open source statistical analysis program that is available free of charge. But before diving into a statistics package there is one important background fundamental that must be covered: data.

-----

## Types of Data

There are two main types of data and it is important to understand the difference between them since that determines appropriate analytical tests.

* Continuous data are integer or decimal numbers and are typically used for counts or measures -- like a person's weight, a tree's height, or a car's speed. Continuous data are measured with scales that have equal divisions so the difference between any two values can be calculated. Because continuous data include characteristics like means and standard deviations, they are analyzed using parametric tests.

* Categorical data group observations into a limited number of categories; for example, type of pet (cat, dog, bird, etc.) or place of residence (Arizona, California, etc.). One common type of categorical data is generated with an "agree-disagree" type of scale, like "I enjoy reading: Strongly Agree : Agree : Neutral : Disagree : Strongly Disagree." Because categorical data do not have characteristics like means or standard deviations, they are analyzed using nonparametric tests.

-----

## The R Command Line

All _R_ commands are entered from a "Command Line" environment. Many students find this a bit challenging at first but once they learn some foundational concepts the command line becomes easy and fast to use. This is an explanation for the _R_ script in the box below.

### Demonstration: The R Command Line

* Line 1: This is a comment that is used to record notes in a script. In _R_, all comments start with a hash-mark (#) and everything after that symbol is ignored. Comments are used frequently in scripts presented in this manual in order to explain what the script is doing. Good programmers comment liberally so team members can easily figure out what they did.

* Line 2: Calculate the value of 3+5.

* Line 3: Calculate the value of "5 + 8 * 2".

* Lines 6-7: These lines create two variables, `MaxScore` and `MinScore`, and then assign values to the variables. You should note two important things about these lines. First, the "assignment" operator is a less than sign followed by a hyphen, making a leftpointing arrow like `<-`. That tells _R_ to store the number on the right side of the arrow operator into the variable named on the left side of the line. Also, keep in mind that capitalization matters with _R_. Thus, the variable named `MaxScore` would be different than a variable named `maxscore`. These lines only store values in variables and nothing gets printed to the screen.

<div class="grsnote">A _variable_ is nothing more than a place in memory to store temporary data. Think of it as a "box" that is used to store something until it is needed later.</div>

* Line 8: The variable `Range` is filled with the result of subtracting `MaxScore` minus `MinScore`.

* Line 9: Entering a variable name, like `Range`, on a line by itself causes the value stored in that variable to be displayed.

* Line 12: In _R_, a list of numbers can be stored in a single variable by using the "combine" function, which is a _c_ followed by a list of the numbers inside a parenthesis. This line creates a variable called `TestScores` and then stores a list of six numbers in that variable.

* Line 13: The contents of the variable `TestScores` is printed to the screen.

```{r 010.02}
# Some basic calculations
3 + 5
5 + 8 * 2

# Using variables
MaxScore <- 57
MinScore <- 22
Range <- MaxScore - MinScore
Range

# The combination function
TestScores <- c(85, 83, 92, 88, 71, 85)
TestScores
```

<div class="grsnote">The _DataCamp_ blocks found throughout this lab manual are designed to "try out" _R_ commands. Click the yellow "Run" button in the lower left corner to execute the entire script and see the results. There are two panels available in the box. Clicking _script.R_ in the top of the box displays the _R_ script and clicking _R Console_ displays the results of executing the script. The commands in the script can be modified and rerun in order to "play around" with _R_. Also, _R_ commands can be entered directly in the *R Console* and executed by tapping the `ENTER` key.</div>

### Guided Practice: The R Command Line

Now is a time to try some of the command line skills demonstrated above. In the following _R_ codebox, calculate these values. 

Notes:

1. In the second line the `^` symbol means "raise to the power of" so that line reads "24 plus the value of 2 raised to the 6th power."

2. In the third line the `sqrt()` function calculates the square root of the number in the parenthesis, or the square root of 9 in this example.

* Set the variable M equal to 15 + ( 37 * 2 )

* Set the variable N equal to 24 + ( 2 ^ 6 )

* Set the variable P equal to 15 - ( sqrt(9) )


```{r 010.03, ex="sc010.01", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r 010.04, ex="sc010.01", type="sample-code"}
# Set M to 15 + ( 37 * 2 )


# Set N to 24 + ( 2 ^ 6 )


# Set P to 15 - ( sqrt(9) )


```


```{r 010.05, ex="sc010.01", type="solution"}

M <- 15 + ( 37 * 2 )

N <- 24 + ( 2 ^ 6 )

P <- 15 - ( sqrt(9) )

```

```{r 010.06, ex="sc010.01", type="sct"}

ex01.01.01neq <- "Check your input for M. It should be 15 + ( 37 * 2 )."
ex01.01.02neq <- "Check your input for N. It should be 24 + ( 2 ^ 6 )."
ex01.01.03neq <- "Check your input for P. It should be 15 - ( sqrt(9) )."

ex() %>% check_object("M") %>% 
  check_equal(incorrect_msg = ex01.01.01neq)

ex() %>% check_object("N") %>% 
  check_equal(incorrect_msg = ex01.01.02neq)

ex() %>% check_object("P") %>% 
  check_equal(incorrect_msg = ex01.01.03neq)

success_msg("Good Work! Now you can use R to calculate math problems!")

```


### Activity 1.1

Start R and calculate the following problems:

* 5 + 8 * 3
* 10 * 2/4
* 5 * 3 * 3 - 20

Include the answers to these three problems in the deliverable document for this lab.

```{r 010.07, ex="act010.01", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r 010.08, ex="act010.01", type="sample-code"}
# Find the value of 5 + 8 * 3

# Find the value of 10 * 2/4

# Find the value of 5 * 3 * 3 - 20



```


-----

## Data Frames

A _data frame_ is a collection of data generated during a research project. An example data frame that is easy to understand would be a spreadsheet that contains the times recorded for a race. _R_ comes configured with a 103 built-in data frames used for training and the _R_ script below is an introduction to one of the data frames used in several of the labs in this manual: [mtcars].

### Demonstration: Data Frames

* Line 2: Entering the name of the data frame, [mtcars], on a line by itself causes _R_ to print the contents of the entire the data frame to the screen. Since [mtcars] is rather small it is fine to print it to the screen, but some data frames have hundreds of lines and that may cause the screen to "scroll" for some time before the end of the data frame is reached.

* Line 3: This prints the structure of the [mtcars] data frame. The result shows that this is a _data.frame_ and has 32 observations (that is how many cars are in the dataset) of 11 variables (things like _mpg_). Also the structure command displays the type of data that are in the dataset. For example, all 11 variables are of the "number" type. The `str` function is frequently used to better understand a data frame. The structure for all of the data frames used in this book is contained in the appendix.

* Line 4: This line prints the maximum _mpg_ value for the [mtcars] data frame. Note that the specific vector desired is indicated by both the data frame name and the vector, separated by a dollar sign, like _mtcars$mpg_ on this line.

_If some of the lines in the result are too long to fit on one row in the R Console they will wrap around._


```{r 010.09}

# Exploring the Motor Trend Cars Data Frame
mtcars
str(mtcars)
max(mtcars$mpg)

```

### Guided Practice: Data Frames

In the following _R_ codebox, explore the [airquality] data frame. 

* Set MaxWind to the maximum value of _Wind_

* Set MaxTemp to the maximum value of _Temp_

* Set MinOzone to the minimum value of _Ozone_

```{r 010.10, ex="sc010.02", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r 010.11, ex="sc010.02", type="sample-code"}
# Set max.wind to the maximum value of Wind


# Set max.temp to the maximum value of Temp


# Set min.ozone to the minimum value of Ozone


```


```{r 010.12, ex="sc010.02", type="solution"}

MaxWind <- max(airquality$Wind)

MaxTemp <- max(airquality$Temp)

MinWind <- min(airquality$Wind)

MinTemp <- min(airquality$Temp)

```

```{r 010.13, ex="sc010.02", type="sct"}

ex01.02.01neq <- "Check your input for MaxWind. It should be max(airquality$Wind)."
ex01.02.02neq <- "Check your input for MaxTemp. It should be max(airquality$Temp)."
ex01.02.03neq <- "Check your input for MinWind. It should be min(airquality$Wind)."
ex01.02.04neq <- "Check your input for MinTemp. It should be min(airquality$Temp)."

ex() %>% check_object("MaxWind") %>% 
  check_equal(incorrect_msg = ex01.02.01neq)
ex() %>% check_object("MaxTemp") %>% 
  check_equal(incorrect_msg = ex01.02.02neq)
ex() %>% check_object("MinWind") %>% 
  check_equal(incorrect_msg = ex01.02.03neq)
ex() %>% check_object("MinTemp") %>% 
  check_equal(incorrect_msg = ex01.02.04neq)

success_msg("Perfect! Thanks for exploring this built-in data frame!")

```


### Activity 1.2

Using the [cafe] data, find the maximum and minimum values for the _bill_ variable and include those values in the deliverable document for this lab.

```{r 010.14, ex="act010.02", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r 010.15, ex="act010.02", type="sample-code"}
# Find the maximum value of bill.

# Find the minimum value of bill.



```

-----

## Deliverable

Complete the activities in this lab and consolidate the responses into a single document. Name the document with your name and "Lab 1," like "George Self Lab 1" and submit that document for grade.
