# Visualizing Data

```{r include = FALSE}

# Load some libraries
library(knitr)
library(kableExtra) # For building pretty tables
library(RColorBrewer) # For enhanced color palettes
library(here)

# Set the rproj home directory
here::here()

# Set some knitr options
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  strip.white = TRUE,
  comment = "#>",
  out.width = "65%",
  message=FALSE,
  warnings=FALSE
)

# Turn on the DataCamp Light interactive tools
tutorial::go_interactive()

```


## Introduction

It is often very useful to visualize data frames rather than simply observe a table. Visualizations can make subtle relationships evident, and those relationships could be missed by just looking at a data table. Earlier tutorials described visualization techniques for descriptive and frequency data and this tutorial applies the same consideration to quantitative data.

-----

## Histogram

A histogram is a graph that shows the disribution of data that are continuous in nature, for example, age or height. A histogram resembles a bar chart but there is an important difference: a histogram is used for continuous data while a bar chart is used for categorical data. To emphasize that difference, histograms are normally drawn with no space between the bars (the data are continuous along the entire x-axis) while bar charts are normally drawn with a small space between bars (the data are categorical along the x-axis). As an example of a histogram, the next figure shows the New York city high temperature from May to September 1973 from the [_airquality_](/tutorials/9999-data-dict/index.html#airquality) data frame.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

mypalette <- brewer.pal(9,"PuOr")

hist(airquality$Temp,
     main = "NY Temperature",
     xlab = "Temperature",
     ylab = "Frequency",
     #labels = TRUE, # Add labels to each bar
     breaks = 12,
     freq = TRUE,
     col = mypalette
     )

```


Notice that there is not a separate bar for each temperature; rather, _R_ has clustered five temperatures into a single bar. Thus, there is a bar that combines the temperatures 70-74 and not separate bars for each of those temperatures. This histogram helps researchers determine if temperature is normally distributed, that is, does it display a typical "bell" curve. It should be fairly obvious from this graph that there are more temperatures around the 80° level than at either extreme so this histogram does indicate that the data are normally distributed.

As another example, the following figure shows a histogram for the body temperature of a beaver recorded every 10 minutes over the course of several hours, as found in the [beaver1](/tutorials/9999-data-dict/index.html#beaver) data frame. This histogram shows normally-distributed data though there is an outlier at a temperature of 37.6°.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

mypalette <- brewer.pal(10,"BrBG")

hist(beaver1$temp,
     main = "Beaver Body Temperature",
     xlab = "Temperature",
     ylab = "Frequency",
     #labels = TRUE, # Add labels to each bar
     breaks = 9,
     freq = TRUE,
     col = mypalette
)

```


Histograms can also indicate data that are skewed and this would be important to researchers during a project's exploratory phase. Consider, for example, the following figure which is the shape of petroleum rock samples in the [_rock_](/tutorials/9999-data-dict/index.html#rock) data frame. While this histogram still indicates a normal distribution with levels falling off from a peak, there is a longer "tail" to the right so the data have a positive skew.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

mypalette <- brewer.pal(9,"RdYlGn")

hist(rock$shape,
     main = "Rock Shape",
     xlab = "Peri / sq root area",
     ylab = "Frequency",
     #labels = TRUE, # Add labels to each bar
     breaks = 12,
     freq = TRUE,
     col = mypalette
)

```


Finally, consider the histogram in the following figure, which is taken from the _peri_ vector in the _rock_ data frame. This shows a bi-modal distribution where there are two clear peaks in the data. It is critical for researchers to know that the data are bi-modal before they begin analyzing it since having two modes can cause certain statistical tests to fail.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

mypalette <- brewer.pal(10,"RdBu")

hist(rock$peri,
     main = "Rock Perimeter",
     xlab = "Perimeter",
     ylab = "Frequency",
     #labels = TRUE, # Add labels to each bar
     breaks = 16,
     freq = TRUE,
     col = mypalette
)

```

### Demonstration: Histogram

The following script creates an example histogram.

* Line 2: This is the beginning of the histogram function (it ends on Line 8). For this histogram the _Speed_ variable from the [_morley_](/tutorials/9999-data-dict/index.html#morley) data frame is specified as the data source for the histogram.
* Lines 3-5: This creates the main title of the histogram along with the labels for the x-axis and y-axis.
* Line 6: To create a histogram, _R_ analyzes the values contained in a variable and creates "bins" for those values. That means that many of the continuous values will be grouped into a single bin for analysis. The "breaks" parameter tells _R_ how many breaks to allow in the variable. In this case, eight breaks are specified, which would create nine bins. _R_ will analyze the data and use the "breaks" parameter as a "suggestion" and will only use that number of breaks if it makes sense for the data being graphed. Often, changing the number of breaks by just one or two will not change the histogram produced so researchers should play around with the "breaks" number to get the best possible representation of the data.
* Line 7: This specifies that 10 colors will be used from the "cm.color" palette to shade the various bars in the histogram. Researchers need to experiment a bit with the color palette and number of colors to get the best result; however, "cm.color" along with the number of bars in the histogram seems to work well for this histogram. (Note: setting colors on graphs was introduced in [Visualizing Descriptives](/tutorials/0040-boxplot/Color).) 


```{r}
# Histogram
hist(morley$Speed,
  main = "Morley's Experiment",
  xlab = "Speed",
  ylab = "Frequency",
  breaks = 8,
  col = cm.colors(10)
)

```

<div class="grsnote">
The DataCamp interface generates graphics in a _Plots_ tab but because of the size of the interface those plots are "squished" and impossible to read. Click the double-headed arrow button on the _Plots_ tab to open the graph in a larger window for evaluation and copying to a document. If the graphic does not open in a larger window then temporarily pause the browser's pop-up blocker.
</div>

### Skill Check: Histogram

Using the [_USJudgeRatings_](/tutorials/9999-data-dict/index.html#usjudgeratings) data frame, create a histogram of the _CONT_ vector. The histogram should have these characteristics:

* Main title: _US Judge Ratings_
* X-axis title: _Integrity Rating_
* Y-axis title: _Frequency_
* Breaks: 11
* Colors: cm.colors(8)

```{r ex="ex08.01", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r ex="ex08.01", type="sample-code"}
# Create the histogram as specified above.


```

```{r ex="ex08.01", type="solution"}

hist(USJudgeRatings$INTG,
  main = "US Judge Ratings",
  xlab = "Integrity Rating",
  ylab = "Frequency",
  breaks = 11,
  col = cm.colors(8)
)

```

```{r ex="ex08.01", type="sct"}

# These are numbered by the exercise number and a sequence within the exercise.
# The sequence number is not necessary if there is only one check.
ex11neq <- "Check the histogram settings to be sure they match the requirement."
ex12neq <- "Check that the histogram is using the USJudgeRatings$INTG vector."
ex13neq <- "Check that main is US Judge Ratings."
ex14neq <- "Check that xlab is Integrity Rating."
ex15neq <- "Check that ylab is Frequency."
ex16neq <- "Check that breaks is 11."
ex17neq <- "Check that col is cm.colors(8)."

ex() %>%
  check_function("hist") %>%
  check_result() %>%
  check_equal(incorrect_msg = ex11neq)

state1 <- ex() %>% check_function("hist")
state1 %>%
  check_arg("x") %>%
  check_equal(incorrect_msg = ex12neq)
state1 %>%
  check_arg("main") %>%
  check_equal(incorrect_msg = ex13neq)
state1 %>%
  check_arg("xlab") %>%
  check_equal(incorrect_msg = ex14neq)
state1 %>%
  check_arg("ylab") %>%
  check_equal(incorrect_msg = ex15neq)
state1 %>%
  check_arg("breaks") %>%
  check_equal(incorrect_msg = ex16neq)
state1 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex17neq)


success_msg("Good Job! Histograms are very important tools for data analysis.")

```

### Activity 8.1: Histogram

Using the _cafe_ data frame, create a histogram of _age_. The histogram should:

1. Have a title of "Activity 8.1: Histogram"
2. The x-axis should have this label: "Age"
3. The y-axis should have this label: "Frequency"
4. Use 10 breaks
5. Use the cm color palette

Copy/paste the histogram in the deliverable document for this lab.

```{r ex="act08.01", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r ex="act08.01", type="sample-code"}
# Using the cafe data frame, create a histogram of age that matches the listed specifications.



```

-----

## Density Plot

A density plot provides the same information as a histogram but it is smoothed out so it is easier to read. Here is the same NY City temperature data from an earlier histogram but drawn as a density plot.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(density(airquality$Temp),
     main = "NY Temperature",
     xlab = "Temperature",
     ylab = "Density",
     lwd = 2,
     bty = "l",
     col = "seagreen4"
)

```

Given the previous plot it is natural to wonder, "What is density?" This is a calculated value such that the total area under the curve is assumed to be one and then each point along the x-axis is calculated to contribute the correct proportional amount to that total density. For many purposes it is adequate to just consider a density plot to be a smoothed histogram. As just one other example, following is a density plot of the bi-modal histogram presented earlier. The density plot makes the bi-modal nature of the data very evident.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(density(rock$peri),
     main = "Rock Perimeter",
     xlab = "Perimeter",
     ylab = "Frequency",
     lwd = 2,
     bty = "l",
     col = "darkorchid4"
)

```

The following script creates a density plot of the _speed_ variable in the _morley_ data frame. The lines are very similar to the last script except _R_ plots the density of the data in Line 2 rather than just a count of the various bins.

```{r}

# Density Plot
plot(density(morley$Speed),
  main = "Morley's Experiment",
  xlab = "Speed",
  ylab = "Density",
  lwd = 2,
  col = "blue4"
)

```

### Demonstration: Density Plot

The following script creates a density plot for the _Rape_ vector in the [_USArrests_](/tutorials/9999-data-dict/index.html#usarrests) data frame. 

* Line 2: Create a density plot for USArrests$Rape.
* Line 3: The main title for the graph.
* Line 4: The title for the x-axis.
* Line 5: The title for the y-axis.
* Line 6: The type of line to draw. "Lwd 2" is a relitively thick line that is easy to see on a graph.
* Line 7: The color for this graph is "magenta 4"

```{r tut=TRUE}
# Density plot of US Arrests for rape.
plot(density(USArrests$Rape),
  main = "US Rape Arrests",
  xlab = "Arrests",
  ylab = "Density",
  lwd = 2,
  col = "magenta4"
)

```

### Skill Check: Density Plot

Using the _USJudgeRatings_ data frame, create a density plot of the _CONT_ vector. The plot should have these characteristics:

* Main title: _US Judge Ratings_
* X-axis title: _Integrity Rating_
* Y-axis title: _Frequency_
* Lwd: 2
* Col: blue4

```{r ex="ex08.02", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r ex="ex08.02", type="sample-code"}
# Create the density plot as specified above.


```

```{r ex="ex08.02", type="solution"}

plot(density(USJudgeRatings$INTG),
  main = "US Judge Ratings",
  xlab = "Integrity Rating",
  ylab = "Frequency",
  lwd = 2,
  col = "blue4"
)

```

```{r ex="ex08.02", type="sct"}

# These are numbered by the exercise number and a sequence within the exercise.
# The sequence number is not necessary if there is only one check.
ex21neq <- "Check the density plot settings to be sure they match the requirement."
ex22neq <- "Check that the density plot is using the USJudgeRatings$INTG vector."
ex23neq <- "Check that main is US Judge Ratings."
ex24neq <- "Check that xlab is Integrity Rating."
ex25neq <- "Check that ylab is Frequency."
ex26neq <- "Check that lwd is 2."
ex27neq <- "Check that col is blue4."

ex() %>%
  check_function("plot") %>%
  check_result() %>%
  check_equal(incorrect_msg = ex21neq)

state21 <- ex() %>% check_function("density")
state1 %>%
  check_arg("x") %>%
  check_equal(incorrect_msg = ex22neq)

state22 <- ex() %>% check_function("plot")
state22 %>%
  check_arg("main") %>%
  check_equal(incorrect_msg = ex23neq)
state22 %>%
  check_arg("xlab") %>%
  check_equal(incorrect_msg = ex24neq)
state22 %>%
  check_arg("ylab") %>%
  check_equal(incorrect_msg = ex25neq)
state22 %>%
  check_arg("lwd") %>%
  check_equal(incorrect_msg = ex26neq)
state22 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex27neq)


success_msg("Good Job! Density plots are very important tools for data analysis.")

```


### Activity 8.2: Density Plot

Using the _cafe_ data frame, create a density plot of _age_. The plot should:

1. Have a title of "Activity 8.2: Density Plot"
2. The x-axis should have this label: "Age"
3. The y-axis should have this label: "Density"
4. Use a line width of 2
5. Use _cadetblue4_ color

Copy/paste the density plot in the deliverable document for this lab.

```{r ex="act08.02", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r ex="act08.02", type="sample-code"}
# Using the cafe data frame, create a density plot of age that matches the listed specifications.



```

-----

## Line Graph

Line graphs display the frequency of some value in a linear form that makes trend detection easier. These types of graphs are especially useful with what is called "time series" data; that is, data that are gathered over a long period of time. As an example, consider the following from the [_airmiles_](/tutorials/9999-data-dict/index.html#airmiles) dataset which charts the number of passenger miles on US commercial airlines from 1937 to 1960.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(airmiles,
     main = "Passenger Air Miles",
     xlab = "Year",
     ylab = "Air Miles",
     type="l",
     lwd = 3,
     col = "olivedrab4"
     )

```

The following figure shows the number of accidental deaths in the United States from 1973 until 1979 by month taken from the [_USAccDeaths_](/tutorials/9999-data-dict/index.html#usaccdeaths) data frame. This line graph clearly shows a seasonal difference where there are more accidental deaths in the summer months than winter and detecting seasonal variation is one of the strengths of a line graph.

```{r fig.width=6,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(USAccDeaths,
     main = "US Accidental Deaths",
     xlab = "Year",
     ylab = "Number of Deaths",
     type="l",
     lwd = 2,
     col = "darkgoldenrod4"
)

```

As one final example, the following figure shows the approval rating for US Presidents from 1945 until 1975 taken from the [_presidents_](/tutorials/9999-data-dict/index.html#presidents) data frame. This line graph shows a very high approval rating in 1945 (Roosevelt at the end of WWII) with dips in the early 1950’s (Truman and the Korean Conflict) and about 1974 (Watergate and Nixon’s resignation). Notice that there are two gaps in the line (late 1940’s and 1973) caused by missing data in the data frame.

```{r fig.width=6,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(presidents,
     main = "US Presidential Approval",
     xlab = "Year",
     ylab = "Rating",
     type="l",
     lwd = 2,
     col = "brown4"
)

```

### Demonstration: Line Graph

Line graphs like those shown above are very useful for detecting a change in some variable, especially over time. The following script creates a line graph of the monthly number of lung disease deaths in the United Kingdom between 1974 and 1979 as found in the [_ldeaths_](/tutorials/9999-data-dict/index.html#ldeaths) data frame. It shows a very interesting cycle where there are consistently more deaths in the winter months than the summer months.

* Line 2: The `plot` function is called for the _ldeaths_ data frame. By itself, `plot` will create a scatter plot but Line 6 in this script specifies a line graph.
* Lines 3-5: This creates the main title of the line graph along with the labels for the x-axis and y-axis.
* Line 6: This specifies the type of line graph wanted. The possible values are: "p" for points, "l" for lines, "b" for both, "c" for the lines part alone of "b", "o" for both ’overplotted,’ "h" for 'histogram-like' (or 'high-density') vertical lines, "s" for stair steps, "S" (that is a capital "S") for other steps, "n" for no plotting. Researchers could quickly try several different settings to determine the best way to present the data and students working through this tutorial can change the value on Line 6 to see the different types of plots.
* Line 7: The _lwd_ parameter sets the width of the line. The default value is 1 so this line specifies a double-width line in order to make it easier to see.
* Line 8: This sets the color of the line to red in order to set it off from the axis and text.

```{r}
# Line Graph
plot(ldeaths,
  main = "Lung Disease Deaths",
  xlab = "Month",
  ylab = "Number",
  type="l",
  lwd = 2,
  col = "red"
)

```

The above graph shows an interesting cyclic variation in the number of deaths over the course of a year. The years listed on the x-axis indicate January of that year. The number of deaths seems to spike high over the winther months and hit a low in about August or September. This type of cycle is very easy to see on this plot but would be difficult to detect in a table full of numbers.

### Skill Check: Line Graph

Using the [_UKgas_](/tutorials/9999-data-dict/index.html#ukgas) data frame, create a line graph that meets this specification:

* plot: UKgas
* main: UK Quarterly Gas Consumption
* xlab: Year
* ylab: Millions of Therms
* type: l
* lwd: 2
* col: lightskyblue

```{r ex="ex08.03", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r ex="ex08.03", type="sample-code"}
# Create a line graph using the specifications listed above.


```

```{r ex="ex08.03", type="solution"}

# Line Graph
plot(UKgas,
  main = "UK Quarterly Gas Consumption",
  xlab = "Year",
  ylab = "Millions of Therms",
  type="l",
  lwd = 2,
  col = "lightskyblue"
)

```

```{r ex="ex08.03", type="sct"}

# These are numbered by the exercise number and a sequence within the exercise.
# The sequence number is not necessary if there is only one check.
ex31neq <- "Check the line graph settings to be sure they match the requirement."
ex32neq <- "Check that the line graph is using the UKgas vector."
ex33neq <- "Check that main is UK Quarterly Gas Consumption."
ex34neq <- "Check that xlab is Year."
ex35neq <- "Check that ylab is Millions of Therms."
ex36neq <- "Check that type is l (that is a lower-case letter L)."
ex37neq <- "Check that lwd is 2."
ex38neq <- "Check that col is lightskyblue."

ex() %>%
  check_function("plot") %>%
  check_result() %>%
  check_equal(incorrect_msg = ex31neq)

state3 <- ex() %>% check_function("plot")
state3 %>%
  check_arg("x") %>%
  check_equal(incorrect_msg = ex32neq)
state3 %>%
  check_arg("main") %>%
  check_equal(incorrect_msg = ex33neq)
state3 %>%
  check_arg("xlab") %>%
  check_equal(incorrect_msg = ex34neq)
state3 %>%
  check_arg("ylab") %>%
  check_equal(incorrect_msg = ex35neq)
state3 %>%
  check_arg("type") %>%
  check_equal(incorrect_msg = ex36neq)
state3 %>%
  check_arg("lwd") %>%
  check_equal(incorrect_msg = ex37neq)
state3 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex38neq)


success_msg("Perfect! A line graph can reveal interesting trends over time.")

```


### Activity 8.3: Line Graph

Create a line graph of the _lynx_ data frame. The plot should:

1. Have a title of "Activity 8.3: Line Graph"
2. The x-axis should have this label: "Year"
3. The y-axis should have this label: "Number"
4. Use a plot type of "l" (this is the letter "l", not the number 1)
5. Use a line width of 2
6. Use the color _firebrick4_

Copy/paste the line graph in the deliverable document for this lab.

```{r ex="act08.03", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r ex="act08.03", type="sample-code"}
# Create a line graph of the lynx data frame that matches the listed specifications.



```

-----

## Plot

Plots (often called "scatter plots") are used to show how two different variables are related. Scatter plots are often used in connection with correlation where they visually indicate the correlation between two variables. For example, The following figure is the plot of stopping distance vs. speed from the [_cars_](/tutorials/9999-data-dict/index.html#cars) data frame.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(cars,
     main = "Stopping Distances",
     xlab = "Speed",
     ylab = "Stopping Distance",
     pch = 21,
     bg = "cyan4",
     col = "black"
     )
```

The previous figure shows that as a car's speed increases the stopping distance also increases, which is exactly what would be expected. (Note: this data were gathered on cars in the 1920s.) Often, a line of best fit is included with a plot to better visualize the relationship between the two variables, as illustrated in the following figure.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(cars,
     main = "Stopping Distances",
     xlab = "Speed",
     ylab = "Stopping Distance",
     pch = 21,
     bg = "cyan4",
     col = "black"
)
abline(lm(cars$dist ~ cars$speed),
       col = "darkred",
       lwd = 2
       )

```

As a second example, The following figure is a plot of the eruption time for Old Faithful geyser, taken from the [_faithful_](/tutorials/9999-data-dict/index.html#faithful) data frame, as a function of the waiting time between eruptions.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

plot(faithful$eruptions ~ faithful$waiting,
     main = "Old Faithful Eruptions",
     xlab = "Waiting Time (Min)",
     ylab = "Eruption Time (Min)",
     pch = 21,
     bg = "peachpuff3",
     col = "black"
)

```

The previous figure shows that as the time between eruptions increases the time that the eruption lasts also increases. Notice that this scatter plot also suggests that the data are bi-modal since there are two clusters of points and a researcher would want to explore that matter before doing much else with the data.

### Demonstration: Plot

The following script creates a simple plot using the [_swiss_](/tutorials/9999-data-dict/index.html#swiss) data frame.

* Line 2: This starts the `plot` function. Like many _R_ functions, `plot` requires a formula input in the form of y ~ x where the dependent variable (the y-axis) is first in the formula and the independent variable (the x-axis) is second. For this plot, _Education_ is the independent variable and will be on the x-axis while _Fertility_ is the dependent variable on the y-axis. The researcher was answering the question "Does education level affect the number of children people have?"
* Lines 3-5: This creates the main title and the labels for the x-axis and y-axis.
* Line 6: The _pch_ attribute is the type of point used on the graph. _R_ has a number of pre-defined types of points and for this graph type number 21 is selected. That is a hollow circle where both the internal color and the circle colors can be specified.
* Line 7: The fill color of the circle is specified using the _bg_ attribute. This this case it is a shade of chartreuse (a yellow-green hue). 
* Line 8: The color of the circle is specified using the _col_ attribute.


```{r}
# Simple Plot
plot(swiss$Fertility ~ swiss$Education,
  main = "Swiss Indicators",
  xlab = "Education",
  ylab = "Fertility",
  pch = 21,
  bg = "chartreuse3",
  col = "black"
)

```

The following script creates a plot with a line of best fit.

* Lines 2-9: This demonstration plots two vectors from the _attitude_ data frame, _rating_ is the independent variable on the x-axis while _complaints_ is the dependent variable on the y-axis. The researcher was answering the question "Do employees with higher overall ratings handle complaints better?" The other  parameters of this `plot` function are defined for the previous plot and are not further discussed here.
* Line 10: This starts the `abline` function that ends on line 13. `Abline` draws a line "from point A to point B" (which is why the function is named "abline") on an existing plot. In this case, the line to be drawn is calculated with the `lm` function, which calculates the slope and y-intercept of the line of best fit for the two specified vectors. The `lm` function was introduced in the tutorial about [Correlation](/tutorials/0070-correlation).
* Lines 11-12: These are the parameters for the line of best fit as drawn on the plot.

```{r}

# Line of Best Fit
plot(attitude$complaints ~ attitude$rating,
  main = "Attitude Data",
  xlab = "Overall Rating",
  ylab = "Complaints",
  pch = 21,
  bg = "khaki2",
  col = "black"
)
abline(lm(attitude$complaints ~ attitude$rating),
  col = "darkseagreen4",
  lwd = 2
)

```


### Skill Check: Plot

Using the _rock_ data frame, create a plot of _area_ (y-axis) as a function of _peri_ (x-axis). The graph should use these specifications:

plot:

* main: Rock Area by Perimeter
* xlab: Perimeter
* ylab: Area
* pch: 21
* bg: goldenrod3
* col: black

abline:

* col: violetred3
* lwd: 2

```{r ex="ex08.04", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r ex="ex08.04", type="sample-code"}
# Create a scatter plot with a line of best fit as specified above.


```

```{r ex="ex08.04", type="solution"}

# Line of Best Fit
plot(rock$area ~ rock$peri,
  main = "Rock Area by Perimeter",
  xlab = "Perimeter",
  ylab = "Area",
  pch = 21,
  bg = "goldenrod3",
  col = "black"
)
abline(lm(rock$area ~ rock$peri),
  col = "violetred3",
  lwd = 2
)

```

```{r ex="ex08.04", type="sct"}

ex401neq <- "Check that the plot formula is rock$area ~ rock$peri."
ex402neq <- "Check that main is Rock Area by Perimeter."
ex403neq <- "Check that xlab is Perimeter."
ex404neq <- "Check that ylab is Area."
ex405neq <- "Check that pch is 21."
ex406neq <- "Check that bg is goldenrod3."
ex407neq <- "Check that col is black."

ex408neq <- "Check that the lm formula is rock$area ~ rock$peri."

ex409neq <- "Check that col is violetred3."
ex410neq <- "Check that lwd is 2."

state41 <- ex() %>% check_function("plot")
state41 %>%
  check_arg("x") %>%
  check_equal(incorrect_msg = ex401neq)
state41 %>%
  check_arg("main") %>%
  check_equal(incorrect_msg = ex402neq)
state41 %>%
  check_arg("xlab") %>%
  check_equal(incorrect_msg = ex403neq)
state41 %>%
  check_arg("ylab") %>%
  check_equal(incorrect_msg = ex404neq)
state41 %>%
  check_arg("pch") %>%
  check_equal(incorrect_msg = ex405neq)
state41 %>%
  check_arg("bg") %>%
  check_equal(incorrect_msg = ex406neq)
state41 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex407neq)

state42 <- ex() %>% check_function("lm")
state42 %>%
  check_arg("formula") %>%
  check_equal(incorrect_msg = ex408neq)


state43 <- ex() %>% check_function("abline")
state43 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex409neq)
state43 %>%
  check_arg("lwd") %>%
  check_equal(incorrect_msg = ex410neq)

success_msg("Perfect! A scatterplot is one of the most common graphs used in research.")

```

### Activity 8.4: Line of Best Fit

Using the _cafe_ data frame, create a simple plot with _length_ on the x-axis, _bill_ on the y-axis, and a line of best fit to see if there is a relationship between the length of the meal and the bill. The plot should:

1. Have a title of "Activity 8.4: Line of Best Fit"
2. The x-axis should have this label: "Length of Meal"
3. The y-axis should have this label: "Bill"
4. Use the color _indianred4_
5. Add a line of best fit with a width of 2 and color of _mediumblue_

Copy/paste the plot in the deliverable document for this lab.

```{r ex="act08.04", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r ex="act08.04", type="sample-code"}
# Using the cafe dataset, create a simple plot of length (x) vs. bill (y) and a line of best fit that matches the listed specifications.



```


-----

## Q-Q Plot

Very often, researchers in the exploratory phase of a project need to know whether a data frame is normally distributed. Creating a histogram or density plot is very helpful in that regard, but a Q-Q ("Quantile-Quantile") Plot is a typical method used to determine if a data frame is normally distributed. The following figure is a Q-Q plot of the New York City temperatures in the _airquality_ data frame.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

qqnorm(airquality$Temp,
       main = "QQ Plot: NY City Temps"
       )
qqline(airquality$Temp,
       col = "Blue",
       lwd = 2
       )
```

An absolutely perfect normal distribution would generate a straight line Q-Q plot and the blue line in the previous figure is ideal. Interpreting a Q-Q plot is more art than science but as long as most values are near the ideal line then the data are considered normally distributed. The following figure is a Q-Q plot for the Old Faithful eruption, as plotted earlier in this tutorial.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

qqnorm(faithful$eruptions,
       main = "QQ Plot: Old Faithful"
)
qqline(faithful$eruptions,
       col = "Dark Green",
       lwd = 2
)
```

The above plot shows a typical bi-modal pattern. On the left side of the plot is a reasonably flat area from -3 to -0.5 on the x-axis and then the plot skips upward and creates a second reasonably flat area between 0.5 and 3. Imagine two parallel lines running through the lower and upper parts of the plot and then notice that the green "ideal" line does not get very close to the slope of either of those two lines. This Q-Q plot shows that the data are not normally distributed. 

The following figure shows a curved Q-Q plot that is typical for a data frame that is skewed. In this case, the plotted sunspot data have a heavy positive skew which is indicated by the long "tail" on the left side of the plot. That skew should be verified with a histogram or density plot.

```{r fig.width=4.5,fig.height=4,fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}

qqnorm(sunspots,
       main = "QQ Plot: Sunspots"
)
qqline(sunspots,
       col = "Red",
       lwd = 2
)

```

### Demonstration: Q-Q Plot

The following script creates a Q-Q plot for a normally-distributed data frame.

* Lines 2-4: This executes the `qqnorm` function and passes that function the weight variable from the _chickwts_ data frame. This function draws the Q-Q Plot. The only parameter needed is _main_, which adds the title to the plot.
* Lines 6-9: This executes the `qqline` function and passes that function the weight variable from the _chickwts_ data frame. This function draws the straight line that indicates a perfect Q-Q plot. The only parameters passed to the function in this script is to set the color to blue and the size to 2, in the same way that was found in the scatterplots above.

```{r}
# Q-Q Plot
qqnorm(chickwts$weight,
  main = "QQ Plot: Chick Weights"
)
# Q=Q Line
qqline(chickwts$weight,
  col = "blue",
  lwd = 2
)

```

### Skill Check: Q-Q Plot

Using the _swiss_ data frame, create a Q-Q plot of the _Fertility_ vector. The plot should have a main title of _QQ Plot: Swiss Fertility_ and a red Q-Q line with a width of 2.

```{r ex="ex08.05", type="pre-exercise-code"}

# No pre-exercise code for this exercise

```

```{r ex="ex08.05", type="sample-code"}
# Create a Q-Q plot of the Fertility vector in the swiss data frame using the requirements listed above.


```

```{r ex="ex08.05", type="solution"}

qqnorm(swiss$Fertility,
  main = "QQ Plot: Swiss Fertility"
)
qqline(swiss$Fertility,
  col = "red",
  lwd = 2
)

```

```{r ex="ex08.05", type="sct"}

ex51neq <- "Check that the qqnorm function is for swiss$Fertility (the capital F on Fertility is important)."
ex52neq <- "Check that main is QQ Plot: Swiss Fertility."
ex53neq <- "Check that the qqline function is for swiss$Fertility (the capital F on Fertility is important)."
ex54neq <- "Check that col is red."
ex55neq <- "Check that lwd is 2."

state51 <- ex() %>% check_function("qqnorm")
state51 %>%
  check_arg("y") %>%
  check_equal(incorrect_msg = ex51neq)
state51 %>%
  check_arg("main") %>%
  check_equal(incorrect_msg = ex52neq)

state52 <- ex() %>% check_function("qqline")
state52 %>%
  check_arg("y") %>%
  check_equal(incorrect_msg = ex53neq)
state52 %>%
  check_arg("col") %>%
  check_equal(incorrect_msg = ex54neq)
state52 %>%
  check_arg("lwd") %>%
  check_equal(incorrect_msg = ex55neq)

success_msg("Great Job! Using Q-Q Plot makes it easy to graphically determine if data are normally distributed.")

```

### Activity 8.5: Q-Q Plot

Using the _cafe_ data frame, create a Q-Q plot of _age_. The plot should:

1. Have a title of "Activity 8.5: Q-Q Plot"
2. A Q-Q line should be added with a color of sienna4 and a weight of 2

Copy/paste the plot in the deliverable document for this lab.

```{r ex="act08.05", type="pre-exercise-code"}

cafe <- read.csv('https://labs.basv316.com/cafe.csv')

```

```{r ex="act08.05", type="sample-code"}
# Using the cafe dataset, create a Q-Q plot of age that matches the listed specification.



```


-----

## Deliverable

Complete the activities in this lab and consolidate the responses into a single document. Name the document with your name and "Lab 8," like "George Self Lab 8" and submit that document for grade.
