# (APPENDIX) Appendix {-} 

# About Colors

```{r 200.01, include = FALSE}

# Load some libraries
library(knitr)
library(kableExtra) # For building pretty tables
library(here)

# Set the rproj home directory
here::here()

# Set some knitr options
options(knitr.table.format = "html")
knitr::opts_chunk$set(
  collapse = TRUE,
  strip.white = TRUE,
  comment = "#>",
  out.width = "65%",
  message=FALSE,
  warnings=FALSE
)

# Turn on the DataCamp Light interactive tools
tutorial::go_interactive()

```


Most visualization tasks in _R_ are enhanced by the use of color. Color can make groups stand out or make it easy to quickly determine what a graph is saying.

## Basic Colors

Consider the following plot of the [faithful] data frame.

```{r 200.02, fig.width=4.5,fig.height=4,fig.align="center", tut=FALSE}
plot(faithful,
     main = "Old Faithful Information",
     xlab = "Eruption Time",
     ylab = "Waiting Time",
     pch = 19
     )
```

The dots in the above scatter plot are black and easy enough to see, but _R_ can color the dots in order to make the visualization more appealing and, perhaps, enhance the readability of the report. _R_ has a vocabulary of 657 basic colors (including 99 levels of grey), and those color names can be spcified with any plot. A PDF file, [Rcolor.pdf](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf), is available that lists all of the colors along with a small sample of each. 

As an example, the following code chunk draws the same [faithful] plot displayed above, but the color _firebrick3_ is named on line 6 so the dots are a red color. That color can be changed to any of the 657 colors available so the plot meets the researcher's needs. The color specification in the following script can be changed and the script re-run to explore the available _R_ colors. 

```{r 200.03}
plot(faithful,
     main = "Old Faithful Information",
     xlab = "Eruption Time",
     ylab = "Waiting Time",
     pch = 19,
     col = "firebrick3"
     )

```

When using color with graphics it is important for the researcher to keep two points in mind.

1. First, it is estimated that about 8% of males and 0.5% of females are unable to distinguish between two or more colors, a condition that is often called “color blindness.”

2. Second, if the research is ever printed in a black-and-white form then all color information is lost.

For these two reasons, it is probably best to not rely on color alone to provide information to the reader; rather, color should be used to enhance the understanding of a chart without being a sole source of information for that chart. 

-----

## Color Palette

Often, a visualization will require more than one color to properly display the data. Consider, for example, this bar plot from the [mtcars] data frame.

```{r 200.04, fig.width=4.5, fig.height=4, out.width="40%", fig.align="center", tut=FALSE}

barplot(height = table(mtcars$cyl, mtcars$gear),
  main = "Cars by Gears and Cylinders",
  xlab = "Gears",
  ylab = "Count",
  legend = TRUE,
  beside = TRUE,
  args.legend = list(title = "Cylinders")
)

```

Since no color is specified, the default colors are applied. The shades of grey are useable but there are easy ways to make the plot more pleasing and useful. _R_ includes five color palettes that contain multiple color combinations and those can be used with visualizations that require more than one color. Here are samples from the five primary color palettes.

```{r 200.05, fig.dim=c(2.5,4.0), fig.align="center", echo=FALSE, message=FALSE, warnings=FALSE, tut=FALSE}
d <- as.table(c(1,1,1,1,1,1,1,1,1,1))
par(mfrow = c(5,1), mai = c(0.35,0.35,0.35,0.35), cex.main = 1.5)
barplot(d, col = cm.colors(10), axisnames = FALSE, axes = FALSE, main = "cm (cyan-magenta)")
barplot(d, col = heat.colors(10), axisnames = FALSE, axes = FALSE, main = "heat")
barplot(d, col = rainbow(10), axisnames = FALSE, axes = FALSE, main = "rainbow")
barplot(d, col = terrain.colors(10), axisnames = FALSE, axes = FALSE, main = "terrain")
barplot(d, col = topo.colors(10), axisnames = FALSE, axes = FALSE, main = "topo")

```

The palettes are applied to a visual with the `col=` command. The number of colors to use from the given palette is specified in the parenthesis following the palette name. Here is how each of the palettes are specified in a visualization.

* _col = cm.colors(5)_
* _col = heat.colors(3)_
* _col = rainbow(7)_
* _col = terrain.colors(8)_
* _col = topo.colors(4)_

Adding the _heat_ color palette on the last line of the script for the barplot makes it easier to read. This is an interactive block of code and the palette can be changed to explore these five palettes. It may also be interesting to change the number of colors to see how that changes the actual colors selected.

```{r 200.06}
barplot(height = table(mtcars$cyl, mtcars$gear),
  main = "Cars by Gears and Cylinders",
  xlab = "Gears",
  ylab = "Count",
  legend = TRUE,
  beside = TRUE,
  args.legend = list(title = "Cylinders"),
  col = heat.colors(3)
)

```

-----

## Custom Palette

Designers must be careful about using multiple hues on a single visual display since that creates what is sometimes called “clown’s pants” due to the extreme patchy color scheme. The goal of using color is to make visualizations easier to understand, but plots with too many colors can be distracting and render the data confusing. Instead, it is generally a best practice to use only shades of the same color or gentle gradients from one color to another. For example, the following script produces the bar plot seen in the previous figure but using only shades of blue.

The first line of the script creates a custom palette named `colpal` (for “color palette”), which, essentially, creates color codes for plots. In this case, the function will create the codes for color gradients between blue and white. _Note: While any two colors can be specified for colorRampPalette, only one hue should be selected along with either white or black in order to make plots more usable for readers who are color blind._ The `col=` line sets the color for this plot by using `colpal` and specifying three colors. To explore this capability, change the colors in the `colorRampPalette()` command and the number of colors to select and run the script.

```{r 200.07}

# Clustered Bar Plot With Gradient Colors
colpal <- colorRampPalette(c("blue", "white"))

barplot(height = table(mtcars$cyl, mtcars$gear),
  main = "Cars by Gears and Cylinders",
  xlab = "Gears",
  ylab = "Count",
  legend = TRUE,
  beside = TRUE,
  args.legend = list(title = "Cylinders"),
  col = colpal(3)
)

```

-----

# About The Data Frames



There are a number of data frames used in the lab exercises in this book and this appendix lists basic information about those data frames.

## airmiles

**Passenger Miles on Commercial US Airlines, 1937-1960**. The revenue passenger miles flown by commercial airlines in the United States for each year from 1937 to 1960. This is a time series of length 24; annually, 1937-1960 (from the FAA Statistical Handbook of Aviation).

## airquality

**New York Air Quality Measurements**. Daily air quality measurements in New York, May to September 1973. This is a data frame with 154 observations on 6 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c("Ozone", 
    "Solar.R", 
    "Wind",
    "Temp",
    "Month",
    "Day"
  ),
  Type = c(
    "int",
    "int",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Ozone (ppb)",
    "Solar R (lang)",
    "Wind (mpg)",
    "Temperature (degrees F)",
    "Month (1-12)",
    "Day of month (1-31)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Air Quality") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## attenu

**The Joyner-Boore Attenuation Data**. This data gives peak accelerations measured at various observation stations for 23 earthquakes in California. The data have been used by various workers to estimate the attenuating affect of distance on ground acceleration. This is a data frame with 182 observations on 5 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "event", 
    "mag", 
    "station",
    "dist",
    "accel"
  ),
  Type = c(
    "numeric",
    "numeric",
    "fac",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Event number",
    "Moment magnitude",
    "Station number (117 levels)",
    "Station-hypocenter distance (km)",
    "Peak acceleration"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Joyner-Boore Attenuation Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## attitude

**The Chatterjee-Price Attitude Data**. From a survey of the clerical employees of a large financial organization, the data are aggregated from the questionnaires of the approximately 35 employees for each of 30 (randomly selected) departments. The numbers give the percent proportion of favourable responses to seven questions in each department. This is a data frame with 30 observations on 7 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "ratings", 
    "complaints", 
    "privileges",
    "learning",
    "raises",
    "critical",
    "advance"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Overall rating",
    "Handling of employee complaints",
    "Does not allow special privileges",
    "Opportunity to learn",
    "Raises based on performance",
    "Too critical",
    "Advancement"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Chatterjee-Price Attitude Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## beaver

**Body Temperature Series of Two Beavers**. Reynolds (1994) describes a small part of a study of the long-term temperature dynamics of beaver Castor canadensis in north-central Wisconsin. Body temperature was measured by telemetry every 10 minutes for four females, but data from a one period of less than a day for each of two animals is used there. There are two data frames: beaver1 has 114 rows and 4 columns on body temperature measurements at 10 minute intervals and beaver2 has 100 rows and 4 columns on body temperature measurements at 10 minute intervals.

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "day", 
    "time", 
    "temp",
    "activ"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Day of observation (days since 1990)",
    "Time of observation (0330 for 3:30 am)",
    "Measured body temperature (Celsius)",
    "Activity outside the retreat (0, 1)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Body Temperature Series of Two Beavers") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## cafe

_This is simulated data._ Customers of the Main Street Cafe completed surveys over a one week period. This is a data frame with observations on 13 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "sex", 
    "age", 
    "day",
    "meal", 
    "length",
    "miles", 
    "pref",
    "ptysize", 
    "food",
    "svc",
    "recmd",
    "bill", 
    "tip"
  ),
  Type = c(
    "fac",
    "int",
    "fac",
    "fac",
    "int",
    "int",
    "fac",
    "int",
    "int",
    "int",
    "fac",
    "numeric",
    "int"
  ),
  Description = c(
    "Sex (3 levels: male, female, other)", 
    "Age", 
    "Day (7 levels: Monday, Tuesday, etc.)",
    "Meal (4 levels: breakfast, lunch, dinner, other)", 
    "Length of meal (minutes)",
    "Miles driven to cafe", 
    "Seating preference (2 levels: booth, table)",
    "Number of people in party", 
    "Rating for food (ord levels: 1-5 'stars')",
    "Rating for service (ord levels: 1-5 'stars')",
    "Would recommend to a friend (2 levels: yes, no)",
    "Bill (dollars and cents)",
    "Amount of tip (whole dollars)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Cafe Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## cars

**Speed and Stopping Distances of Cars**. The data give the speed of cars and the distances taken to stop. Note that the data were recorded in the 1920s. This is a data frame with 50 observations on 2 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "speed", 
    "dist"
  ),
  Type = c(
    "numeric",
    "numeric"
  ),
  Description = c(
    "Speed (mpg)",
    "Stopping distance (ft)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Speed and Stopping Distances") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## ChickWeight

**Weight Versus Age of Chicks on Different Diets**. The ChickWeight data frame was generated from an experiment on the effect of diet on early growth of chicks. This is a data frame with 578 observations on 4 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "weight", 
    "Time",
    "Chick",
    "Diet"
  ),
  Type = c(
    "numeric",
    "numeric",
    "ord",
    "fac"
  ),
  Description = c(
    "Weight (grams)",
    "Time (days since birth)",
    "A uniqe identifier for each chick",
    "A factor of 1-4 indicating the diet"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Chick Weights") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## chickwts

**Chicken Weights by Feed Type**. An experiment was conducted to measure and compare the effectiveness of various feed supplements on the growth rate of chickens. This is a data frame with 71 observations on 2 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "weight", 
    "feed"
  ),
  Type = c(
    "numeric",
    "fac"
  ),
  Description = c(
    "Weight (unk units)",
    "Feed type"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Chicken Weights") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## CO2

**Carbon Dioxide Uptake in Grass Plants**. This data set is from an experiment on the cold tolerance of the grass species _Echinochloa crusgalli_. This is a data frame with 84 observations on 5 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Plant", 
    "Type", 
    "Treatment",
    "conc", 
    "uptake"
  ),
  Type = c(
    "ordered",
    "fac",
    "fac",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Unique identifier of each plant (12 levels)", 
    "Origin of the plant (Quebec, Mississippi)", 
    "Type of treatment (nonchilled, chilled)",
    "Ambient CO2 concentratino (mL/L)", 
    "CO2 uptake rate (umol/sq meter/sec)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Carbon Dioxide Uptake in Grass Plants") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## esoph

**Smoking, Alcohol and (O)esophageal Cancer** . Data from a case-control study of (o)esophageal cancer in Ille-et-Vilaine, France. This is a data frame with 88 observations on 5 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "agegp", 
    "alcgp", 
    "tobgp",
    "ncases", 
    "ncontrols"
  ),
  Type = c(
    "ordered",
    "ordered",
    "ordered",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Age group (6 levels)", 
    "Alcohol consumption (4 levels)", 
    "Tobacco consumption (4 levels)",
    "Number of cases", 
    "Number of controls"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Smoking, Alcohol and (O)esophageal Cancer") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## faithful

**Old Faithful Geyser Data**. Waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA. This is a data frame with 272 observations on 2 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "eruptions", 
    "waiting"
  ),
  Type = c(
    "numeric",
    "numeric"
  ),
  Description = c(
    "Eruption time (minutes)",
    "Waiting time to next eruption (minutes)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Old Faithful Geyser Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## infert

**Infertility Study**. This is a matched case-control study. This is a data frame with 248 observations on 8 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "education", 
    "age", 
    "parity",
    "induced", 
    "case",
    "spontaneous",
    "stratum",
    "pooled.stratum"
  ),
  Type = c(
    "factor",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "integer",
    "numeric"
  ),
  Description = c(
    "Education (0=0-5yrs, 1=6-11yrs, 2=12+yrs)", 
    "Age in years", 
    "Parity",
    "number of prior induced abortions (0=0, 1=1, 2= 2+)", 
    "Status (0=control, 1=case)",
    "number of prior spontaneous abortions (0=0, 1=1, 2= 2+)", 
    "Matched Set Number 1-83",
    "Stratum Number 1-63"
  )
)

kable(cor_tbl, 
      "html",
      caption = "") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## iris

**Edgar Anderson’s Iris Data**. This famous (Fisher’s or Anderson’s) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris: Iris setosa, versicolor, and virginica. This is a data frame with 150 observations on 5 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Sepal.Length", 
    "Sepal.Width", 
    "Petal.Length",
    "Petal.Width", 
    "Species"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "factor"
  ),
  Description = c(
    "Length (cm)", 
    "Width (cm)", 
    "Length (cm)",
    "Width (cm)", 
    "Species (3 levels)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## LakeHuron

**Level of Lake Huron 1875-1972**. Annual measurements of the level, in feet, of Lake Huron 1875-1972. This is a time series of length 98.

## ldeaths

**Monthly Deaths from Lung Diseases in the UK**. Monthly deaths recorded from bronchitis, emphysema and asthma in the UK. This is a time series of length 72, 1974-1979.

## longley

**Longley’s Economic Regression Data**. This data set is a macroeconomic data set which provides a well-known example for a highly collinear regression. This is a data frame with 7 variables observed annually from 1947 to 1962 (n = 16):

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "GNP.deflator", 
    "GNP", 
    "Unemployed",
    "Armed.Forces", 
    "Population",
    "Year", 
    "Employed"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "GNP implicit price deflator (1954=100)", 
    "Gross National Product", 
    "Number employed",
    "Number in armed forces", 
    "Population > 14 years of age",
    "Year", 
    "Number employed"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Longley’s Economic Regression Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## lynx

**Annual numbers of lynx trappings for 1821–1934 in Canada**. This is a time series of length 114, 1821-1934.

## morley

**Michelson Speed of Light Data**. A classical data of Michelson (but not this one with Morley) on measurements done in 1879 on the speed of light. The data consists of five experiments, each consisting of 20 consecutive "runs". The response is the speed of light measurement, suitably coded (km/sec, with 299,000 subtracted). This is a data frame with 100 observations on 3 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Expt", 
    "Run", 
    "Speed"
  ),
  Type = c(
    "int",
    "int",
    "int"
  ),
  Description = c(
    "The experiment number, from 1 to 5", 
    "The run number within each experiment", 
    "Speed-of-light measurement"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Speed of Light") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## mtcars

**Motor Trend Car Road Tests**. The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models). This is a data frame with 32 observations on 11 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "mpg", 
    "cyl", 
    "disp",
    "hp", 
    "drat",
    "wt", 
    "qsec",
    "vs", 
    "am",
    "gear",
    "carb"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Miles/(US) gallon", 
    "Number of cylinders", 
    "Displacement (cu. in.)",
    "Gross horsepower", 
    "Rear axle ratio",
    "Weight (1000 lbs)",
    "Quarter-mile time",
    "V8 (0) or straight (1) engine", 
    "Automatic (0) or manual (1) transmission",
    "Number of forward gears",
    "Number of carburetors"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Motor Trend Car Road Tests") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## npk

**Classical N, P, K Factorial Experiment**. A classical N, P, K (nitrogen, phosphate, potassium) factorial experiment on the growth of peas conducted on 6 blocks. Each half of a fractional factorial design confounding the NPK interaction was used on 3 of the plots. This is a data frame with 24 observations on 5 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "block", 
    "N", 
    "P",
    "K", 
    "yield"
  ),
  Type = c(
    "factor",
    "factor",
    "factor",
    "factor",
    "numeric"
  ),
  Description = c(
    "Which block (6 levels)", 
    "Indicator for nitrogen (No, Yes)", 
    "Indicator for phosphate (No, Yes)",
    "Indicator for potassium (No, Yes)", 
    "Yield of peas (pounds/plot)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "N, P, K Experiment") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## PlantGrowth

**Results from an Experiment on Plant Growth**. Results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions. This is a data frame with 30 observations on 2 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "weight", 
    "group" 
  ),
  Type = c(
    "numeric",
    "factor"
  ),
  Description = c(
    "Weight of yield (in)", 
    "Treatment group (3 levels)" 
  )
)

kable(cor_tbl, 
      "html",
      caption = "Plant Growth") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## presidents

**Quarterly Approval Ratings of US Presidents**. The (approximate) quarterly approval rating for the President of the United States from the first quarter of 1945 to the last quarter of 1974. This is a time series of length 120.

## rivers

**Lengths of Major North American Rivers**. This data set gives the lengths (in miles) of 141 "major" rivers in North America, as compiled by the US Geological Survey. This is a vector with 141 observations.

## rock

**Measurements on Petroleum Rock Samples**. This data set contains measurements on 48 rock samples from a petroleum reservoir. This is a data frame with 48 observations on 4 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "area", 
    "peri",
    "shape",
    "perm"
  ),
  Type = c(
    "int",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Area of pores (pixels in 256 X 256)", 
    "Perimeter in pixels",
    "perimeter/square root of area",
    "Permeability in mili-Darcies"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Petroleum Rock Samples") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## sleep

**Student’s Sleep Data**. Data which show the effect of two soporific drugs (increase in hours of sleep compared to control) on 10 patients. This is a data frame with 20 observations on 3 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "extra", 
    "group",
    "ID"
  ),
  Type = c(
    "numeric",
    "fac",
    "fac"
  ),
  Description = c(
    "Increase in hours of sleep",
    "Drug given (2 levels)",
    "Patient ID (10 levels)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Sleep Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## stackloss

**Brownlee's Stack Loss Plant Data**. Operational data of a plant for the oxidation of ammonia to nitric acid. This is a data frame with 21 observations on 4 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Air Flow", 
    "Water Temp",
    "Acid Conc",
    "stack.loss"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Flow of cooling air",
    "Cooling water inlet temperature",
    "Concentration of acid [per 1000, minus 500]",
    "Stack loss"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Sleep Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## state.division

**US State Facts and Figures**. Data sets related to the 50 states of the United States of America. This is a factor giving the state division (New England, Middle Atlantic, South Atlantic, East North Central, East South Central, West North Central, West South Central, Mountain, Pacific).

## sunspots

**Monthly Sunspot Numbers, 1749-1983**. Monthly mean sunspot numbers from 1749 to 1983. Collected at Swiss Federal Observatory, Zurich until 1960, then Tokyo Astronomical Observatory. This is a time series of length 2820.

## swiss

**Swiss Fertility and Socioeconomic Indicators (1888) Data**. Standardized fertility measure and socio-economic indicators for each of 47 French-speaking provinces of Switzerland at about 1888. This is a data frame with 47 observations on 6 variables, each of which is in percent (0-100):

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Fertility", 
    "Agriculture", 
    "Examination",
    "Education", 
    "Catholic",
    "Infant.Mortality"
  ),
  Type = c(
    "numeric",
    "numeric",
    "int",
    "int",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Common fertility measure", 
    "% males involved in agriculture", 
    "% receiving high mark on army exam",
    "% education beyond primary school", 
    "% catholic (as opposed to protestant)",
    "Live births who live less than 1 year"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Swiss Socioeconomic Indicators (1888)") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## Titanic

**Survival of passengers on the Titanic**. This data set provides information on the fate of passengers on the fatal maiden voyage of the ocean liner "Titanic," summarized according to economic status (class), sex, age and survival. This is a 4-dimensional array resulting from cross-tabulating 2201 observations on 4 variables. The variables and their levels are as follows:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Num = c(
    "1", 
    "2", 
    "3",
    "4"
  ),
  Name = c(
    "Class",
    "Sex",
    "Age",
    "Survived"
  ),
  Levels = c(
    "1st, 2nd, 3rd, Crew", 
    "Male, Female", 
    "Child, Adult",
    "No, Yes"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Titanic") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## trees

**Girth, Height and Volume for Black Cherry Trees**. This data set provides measurements of the girth, height and volume of timber in 31 felled black cherry trees. Note that girth is the diameter of the tree (in inches) measured at 4 ft 6 in above the ground. This is a data frame with 31 observations on 3 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Girth", 
    "Height",
    "Volume"
  ),
  Type = c(
    "numeric",
    "numeric",
    "numeric"
  ),
  Description = c(
    "Tree diameter (inches)",
    "Tree height (feet)",
    "Volume of timber (cubic ft)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Black Cherry Tree Data") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## UCB Admissions

**Student Admissions at UC Berkeley**. Aggregate data on applicants to graduate school at Berkeley for the six largest departments in 1973 classified by admission and sex. This is a 3-dimensional array resulting from cross-tabulating 4526 observations on 3 variables. The variables and their levels are as follows:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Num = c(
    "1", 
    "2",
    "3"
  ),
  Name = c(
    "Admit",
    "Gender",
    "Dept"
  ),
  Levels = c(
    "Admitted, Rejected",
    "Male, Female",
    "A, B, C, D, E, F"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Admissions at UC Berkeley") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## UKgas

**UK Quarterly Gas Consumption**. Quarterly UK gas consumption from 1960Q1 to 1986Q4, in millions of therms. This is a time series of length 108.

## USAccDeaths

**Accidental Deaths in the US 1973-1978**. The monthly totals of accidental deaths in the USA. This is a time series of length 72. 

## USArrests

**Violent Crime Rates by US State**. This data set contains statistics about the arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states in 1973. Also given is the percent of the population living in urban areas. This is a data frame with 50 observations on 4 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "Murder", 
    "Assault",
    "UrbanPop",
    "Rape"
  ),
  Type = c(
    "numeric", 
    "integer",
    "integer",
    "numeric"
  ),
  Definition = c(
    "Murder arrests per 100,000",
    "Assault arrests per 100,000",
    "Percent urban population",
    "Rape arrests per 100,000"
  )
)

kable(cor_tbl, 
      "html",
      caption = "US Arrests") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```

## USJudgeRatings

**Lawyers' Ratings of State Judges in the US Superior Court**. This contains lawyers' ratings of state judges in the US Superior Court. This is a data frame with 43 observations on 12 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "CONT", 
    "INTG",
    "DMNR",
    "DILG",
    "CFMG",
    "DECI",
    "PREP",
    "FAMI",
    "ORAL",
    "WRIT",
    "PHYS",
    "RTEN"
  ),
  Type = c(
    "numeric", 
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric"
  ),
  Definition = c(
    "Number of contacts of lawyer with judge",
    "Judicial integrity",
    "Demeanor",
    "Diligence",
    "Case flow managing",
    "Prompt decisions",
    "Preparation for trial",
    "Familiarity with law",
    "Sound oral rulings",
    "Sound written rulings",
    "Physical ability",
    "Worthy of retention"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Judge Ratings") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


## warpbreaks

**The Number of Breaks in Yarn During Weaving**. This data set gives the number of warp breaks per loom, where a loom corresponds to a fixed length of yarn. This is a data frame with 54 observations on 3 variables:

```{r tut=FALSE, echo=FALSE, message=FALSE, warnings=FALSE}
cor_tbl <- data.frame(
  Name = c(
    "breaks", 
    "wool",
    "tension"
  ),
  Type = c(
    "numeric",
    "fac",
    "fac"
  ),
  Description = c(
    "Number of breaks",
    "Type of wool (levels A, B)",
    "Tension on wool (levels L, M, H)"
  )
)

kable(cor_tbl, 
      "html",
      caption = "Breaks in Yarn During Weaving") %>%
  kable_styling(bootstrap_options = c("striped","hover","responsive"),
                full_width = F,
                position = "center")
  # column_spec(1, bold = T, border_right = T)
```


